Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2011-10-18 ms.tex
Further changes according to Jeremy's and Gordana's suggestions

Galaxy clusters have, for many years, been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is
to be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay,
the charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
  which subsequently decays to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
size of a cluster and the volume-filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable to or longer than the Hubble time
\citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them to accumulate. For a
given CR distribution function, the hadronically induced gamma-ray flux is directly proportional to
the CR-to-thermal pressure fraction, $X_\CR=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$. A very
modest $X_{\CR}$ of a few percent implies an observable flux of gamma rays
\citep[e.g.,][]{article:PfrommerEnsslin:2004b}.

@@

The data analysis was performed following the standard VERITAS procedures described in
\citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction
and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for
events with at least two contributing telescope images that pass the following quality selection
criteria: more than four participating pixels in the camera, number of photoelectrons in the image
is larger than 75, and the distance from the image centroid to the center of the camera is less
than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{Defined
as the energy corresponding to the maximum of the product function of the observed spectrum with
the collection area.} of about 220 GeV. In addition, events for which only images from the two
closest-spaced telescopes\footnote{In the array configuration prior to summer 2009, two telescopes 
had a separation of only 35 m.} survive quality selection are rejected, as they introduce an
irreducible high background rate, due to local muons, degrading the instrument sensitivity
\citep{article:MaierKnapp:2007}.

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of 2.4 using data taken on the Crab Nebula during the same epoch. Because the VHE
gamma-ray spectrum for the Coma cluster is expected to be hard, with a photon index of about 2.3
\citep{article:PinzkePfrommer:2010}, these cuts are suitable for the analysis of the Coma cluster
data set.

@@

The ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background
due to CRs misinterpreted as gamma rays (the cuts described above reject more than 99\% of all
CRs). The total number of events in a given region of interest is then compared to the estimated
background from the off-source region scaled by the ratio of the solid angles to produce a final
excess. The VHE gamma-ray significance is then calculated using the method by
\citet{article:LiMa:1983}. Significance skymaps over the VERITAS field of view produced with a
$0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio
contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz
observations \citep{article:BrownRudnick:2010} respectively.

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to a radius of 0.42 square degrees to cover all regions of interest
considered in this work. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows that the VERITAS data
is well fit by a Gaussian with a mean close to zero and a standard deviation within a few percent
of unity.

@@

\section{\Fermi-LAT Analysis and Results}
The Large Area Telescope (LAT) on board the \Fermi Gamma-ray Space Telescope (\Fermi)
has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The
\Fermi-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is thus
very complimentary to the VERITAS observations. \citet{article:Ackermann_etal:2010} reported on the
search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18
months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$
s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the
exposure is increased. In this work an updated analysis which includes additional data taken during
the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \Fermi Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events
calibrated in flight) were used throughout this work. Only photons above 1 GeV are considered, where
the constraints on the flux are most meaningful. The \Fermi-LAT collaboration estimates the
systematic uncertainties at 10 GeV to be around
20\%.\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper
limits at the 99\% confidence level in the energy range 1--30 GeV, assuming both a unresolved,
point-like or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray
emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and
$2.5$.

@@

\section{Discussion and Conclusions}
VERITAS observed the Coma cluster of galaxies for a total of 18.6 hours of high-quality live time
between March and May in 2008. No significant excess of gamma-rays above a threshold of about 220
GeV was detected. Flux upper limits at 99\% confidence level were calculated for the cluster core
and three member galaxies. The core was considered both as a point-like and spatially extended
source. The obtained flux upper limits were used to constrain properties of the cluster.

We employed various approaches to constrain the CR population and magnetic field distribution that
are complementary in their assumptions and hence well suited in assessing underlying Bayesian
priors in the models. (1) We use a simplified ``isobaric CR model'' that is characterized by a
constant CR-to-thermal pressure fraction and has a power-law momentum spectrum.  While this model
is physically not justified a priory, it is simple and widely used in the literature and captures
some aspects of more elaborate models such as (2) the simulation-based analytical approach of
\citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR
distribution spectrally and spatially for a given set of assumptions (namely the physics that was
simulated). It is powerful since it only requires the density profile as input due to the
approximate universality of the CR distribution (when neglecting CR diffusion and streaming which
appears necessary to explain the radio halo bimodality). (3) We finally use a pragmatic
approach which models the CR and magnetic distribution in order to reproduce the observed emission
profile of the Coma radio halo. While this approach is also physically not justified, it is
powerful as it shows what the ``correct'' model has to achieve and can point into the direction of
the relevant physics.

@@

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The
points with error bars represent the ON-source data sample and the filled area is the estimated
background due to cosmic rays. Each bin represents an annulus around the Coma cluster core
position. The data were derived from the ring background model using a $0.2^{\circ}$ sampling
radius.}
\label{fig:thetasq}
\end{figure*}

>>>>

Galaxy clusters have, for many years, been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is
to be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay,
the charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
which subsequently decays to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
size and the volume-filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable to or longer than the Hubble time
\citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them to accumulate. For a
given CR distribution function, the hadronically induced gamma-ray flux is directly proportional to
the CR-to-thermal pressure fraction, $X_\CR=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$. A very
modest $X_{\CR}$ of a few percent implies an observable flux of gamma rays
\citep[e.g.,][]{article:PfrommerEnsslin:2004b}.

@@

The data analysis was performed following the standard VERITAS procedures described in
\citet{inproc:Cogan_etal:2007} and \citet{inproc:Daniel_etal:2007}. Prior to event reconstruction
and selection, all shower images are calibrated and cleaned. Showers are then reconstructed for
events with at least two contributing telescope images that pass the following quality selection
criteria: more than four participating pixels in the camera, number of photoelectrons in the image
is larger than 75, and the distance from the image centroid to the center of the camera is less
than $1.43^{\circ}$. These quality selection criteria impose an energy threshold\footnote{The energy
threshold is defined as the energy corresponding to the maximum of the product function of the
observed spectrum with the collection area. It does not vary significantly for the different source
scenarios and assumed spectral indices reported in this work.} of about 220 GeV. In addition,
events for which only images from the two closest-spaced telescopes\footnote{In the arra
configuration prior to summer 2009, two telescopes  had a separation of only 35 m.} survive quality
selection are rejected, as they introduce an irreducible high background rate, due to local muons, 
degrading the instrument sensitivity \citep{article:MaierKnapp:2007}.

Gamma-ray like events are separated from the CR background by imposing cuts on scaled parameters
\citep{article:Aharonian_etal:1997, article:Krawczynski_etal:2006} calculated from a parametrized
moment analysis of the shower images \citep{inproc:Hillas:1985}. The gamma-hadron separation cuts
used in this analysis were optimized a priori for a weak point source (3\% Crab Nebula flux level)
and a photon index of 2.4 using data taken on the Crab Nebula during the same epoch. Because the VHE
gamma-ray spectrum for the Coma cluster is expected to be hard, with a photon index of about 2.3
\citep{article:PinzkePfrommer:2010}, these cuts are suitable for the analysis of the Coma cluster
data set. It is noted that slightly varying the spectral index ($\pm$ 0.2) does not significantly
impact the cuts used for quality selection cuts and gamma-hadron separation in this work.

@@

The ring background model \citep{article:Aharonian_etal:2001} is used to estimate the background
due to CRs misinterpreted as gamma rays (the cuts described above reject more than 99\% of all
CRs). The total number of events in a given region of interest is then compared to the estimated
background from the off-source region scaled by the ratio of the solid angles to produce a final
excess. The VHE gamma-ray significance is then calculated according to Formula 17 in
\citet{article:LiMa:1983}. Significance skymaps over the VERITAS field of view produced with a
$0.2^{\circ}$ sampling radius are shown in Figure \ref{fig:skymaps} with overlaid X-ray and radio
contours from the ROSAT all-sky survey \citep{article:BrielHenryBohringer:1992}  and GBT 1.4 GHz
observations \citep{article:BrownRudnick:2010} respectively.

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma-rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to 0.42 square degrees to cover both the case of point-like and extended
emission from the core of the Coma cluster. The $\theta^{2}$ distributions for the member galaxies
alos considered in this work are very similar to that in Figure \ref{fig:thetasq} and show no
excess of gamma rays. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a Gaussian
background. A lower bound of zero is imposed on the signal rate to handle the negative excesses
seen for several of the regions of interest. Figure \ref{fig:sigdist} shows the distribution of
significances over the VERITAS skymap, which is well fit by a Gaussian with a mean close to zero and
a standard deviation within a few percent of unity.

@@

\section{\Fermi-LAT Analysis and Results}
The Large Area Telescope (LAT) on board the \Fermi Gamma-ray Space Telescope (\Fermi)
has observed the Coma cluster in all-sky survey mode since its launch in June 2008. The
\Fermi-LAT is sensitive to gamma rays in the 20 MeV to $\sim$300 GeV energy range and is
very complimentary to the VERITAS observations. \citet{article:Ackermann_etal:2010} reported on the
search for gamma-ray emission from thirty-three galaxy clusters in the data from the first 18
months, including the Coma cluster, for which an upper limit of $4.58\times 10^{-9}$ ph. cm$^{-2}$
s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the
exposure is increased. In this work an updated analysis which includes additional data taken during
the period 2010 February 5 to 2011 June 10 is presented as a compliment to the VERITAS results.

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \Fermi Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse were used throughout this work. Only photons above 1 GeV
are considered, where the constraints on the flux are most meaningful. The \Fermi-LAT collaboration
estimates the systematic uncertainties at 10 GeV to be around
20\%.\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper
limits at the 99\% confidence level in the energy range 1--30 GeV, assuming both an unresolved,
point-like or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray
emission is assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and
$2.5$.

@@

\section{Discussion and Conclusions}
We have reported on the observations of the Coma cluster of galaxies in VHE gamma rays with VERITAS
and complementary observations with the \emph{Fermi}-LAT. VERITAS observed the Coma cluster of
galaxies for a total of 18.6 hours of high-quality live time between March and May in 2008. No
significant excess of gamma-rays above an energy threshold of about 220 GeV was detected. The
\emph{Fermi}-LAT has observed the Coma cluster in all-sky survey mode since its launch in June
2008. We have used all data available from launch to June 2011 for an updated analysis over already
published results \citep{article:Ackermann_etal:2010}. Again, no significant excess of gamma rays
was detected. We have used the VERITAS and \emph{Fermi}-LAT data to calculate flux upper limits at
the 99\% confidence level for the cluster core, considered as both a point-like and spatially
extended source, and three member galaxies. The flux upper limits obtained was then used to
constrain properties of the cluster.

We have employed various approaches to constrain the CR population and magnetic field distribution
that are complementary in their assumptions and hence well suited in assessing underlying Bayesian
priors in the models. (1) We used a simplified ``isobaric CR model'' that is characterized by a
constant CR-to-thermal pressure fraction and has a power-law momentum spectrum. While this model
is physically not justified a priory, it is simple and widely used in the literature and captures
some aspects of more elaborate models such as (2) the simulation-based analytical approach of
\citet{article:PinzkePfrommer:2010}. This is a ``first principle approach'' that predicts the CR
distribution spectrally and spatially for a given set of assumptions (namely the physics that was
simulated). It is powerful since it only requires the density profile as input due to the
approximate universality of the CR distribution (when neglecting CR diffusion and streaming which
appears necessary to explain the radio halo bimodality). (3) We finally used a pragmatic
approach which models the CR and magnetic distribution in order to reproduce the observed emission
profile of the Coma radio halo. While this approach is also physically not justified, it is
powerful as it shows what the ``correct'' model has to achieve and can point into the direction of
the relevant physics.

@@

We have also used the flux upper limits obtained with VERITAS to constrain the thermally averaged
product of the total self-annihilation cross section and the relative velocity of the dark matter
particles. Assuming the dark matter is a WIMP and follows the density profile by 
\citet{article:NavarroFrenkWhite:1997} we calculated the limits on $\expval{\sigma v}$ to be on the
order of $10^{-20}$ to $10^{-21}$ cm$^{-3}$ s$^{-1}$ depending on the chosen aperture. These limits
are based on conservative estimates of the astrophysical factor, which does not inlcude any boosting
from clumping or the Sommerfeld effect.

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f2.eps}}
\end{center}
\caption{$\theta^{2}$ distribution from VERITAS observations of the Coma cluster of galaxies. The
points with error bars represent the ON-source data sample and the filled area is the background estimation in the OFF-source regions. Each bin represents an annulus around the Coma cluster core
position. The data were derived from the ring background model using a $0.2^{\circ}$ sampling
radius.}
\label{fig:thetasq}
\end{figure*}
